---
title: "Parallel programming"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "and dealing with large data..."
date: 2024-11-19
---

# Pre-lecture activities

::: callout-important

In advance of class, please

* `future` - this provides a unified parallel framework in R consistent

You can do this by calling

```{r}
#| eval: false
install.packages("future")
```

And load the package using: 

```{r}
library(future)
```


:::

In addition, please read through

- [Strategies for dealing with large data](../../readings/07-strategies-big-data/index.qmd)
- <https://www.futureverse.org/packages-overview.html> (just the `future` R package)


::: callout-note

## How much should I prepare for before class?

You should have `future` installed and be familiar with the three basic functions - `plan()`, `future()`, and `value()`. 

We will learn more about these functions in class. 


:::


# Lecture 

## Acknowledgements

Material for this lecture was borrowed and adopted from

- <https://www.futureverse.org/packages-overview>

## Learning objectives

::: callout-note
### Learning objectives

**At the end of this lesson you will:**

- Understand the basics of parallel computing
- Become familiar with basic functions in the `future` package
- Recognize different file formats to work with large data not locally
- Implement three ways to work with large data: 
  1. "sample and model" 
  2. "chunk and pull"
  3. "push compute to data"

:::

## Slides 

- [Lecture 07: Parallel computing (and dealing with large data)](https://docs.google.com/presentation/d/1jaMlB3SibUmF-rVFnKf3Hz-8_biX4HO80qt6Kx2Wrd0/edit?usp=sharing)


# Class activity 

For the rest of the time in class, we will practice using the `future` package.
There are two tutorials for you to work through own your own developed by 
Henrik Bengtsson from the UseR! 2024 conference: 

- [Lab: Futureverse 1](https://henrikbengtsson.github.io/future-tutorial-user2024/lab1.html)
- [Lab: Futureverse 2](https://henrikbengtsson.github.io/future-tutorial-user2024/lab2.html)

Also, if you would like to try out the three strategies we learned about in 
class today for dealing with large data, try working through the pre-reading: 

- [Strategies for dealing with large data](../../readings/07-strategies-big-data/index.qmd)