---
title: "Parallel programming"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "and dealing with large data..."
date: 2024-11-19
---

# Pre-lecture activities

::: callout-important

In advance of class, please

* `future` - this provides a unified parallel framework in R consistent

You can do this by calling

```{r}
#| eval: false
install.packages("future")
```

And load the package using: 

```{r}
library(future)
```


:::

In addition, please read through

- [Strategies for dealing with large data](../../readings/07-strategies-big-data/index.qmd)
- <https://www.futureverse.org/packages-overview.html> (just the `future` R package)


::: callout-note

## How much should I prepare for before class?

You should have `future` installed and be familiar with the three basic functions - `future()`, `resolved()`, and `value()`. 

We will learn more about these functions in class. 


:::


# Lecture 

## Acknowledgements

Material for this lecture was borrowed and adopted from

- <https://www.futureverse.org/packages-overview>

## Learning objectives

::: callout-note
### Learning objectives

**At the end of this lesson you will:**

- Understand the basics of parallel and distributed computing
- Become familiar with basic functions in the `future` package
- Recognize different file formats to work with large data not locally
- Implement three ways to work with large data: 
  1. "sample and model" 
  2. "chunk and pull"
  3. "push compute to data"


:::



## Slides 

- <add here>


# Post-lecture


## Additional practice

Here are some additional practice questions to help you think about the material discussed.

::: callout-note
### Questions

1. 
:::